On the Synchronization Power of OpenFlow

### 2015-03-05 inband-STN.tex
new intro (and story)

%\usepackage[caption=true,font=footnotesize]{subfig}
%\usepackage{xspace} 	% Guesses whether a space is needed when invoked
\usepackage{cite}
%\usepackage{framed}
\usepackage{framed,color}
\definecolor{shadecolor}{rgb}{0.9,0.9,0.9}

@@

%\title{Distributed Network Programming}
%\title{The Distributed SDN Update Problem:\\Towards Software Transactional Networks}

%\title{On Consistent Updates in Software Defined Networks under Unreliable Control}
%\title{The Case for Reliable Software Transactional Networking}

%\title{A Distributed SDN Control Plane for Concurrent Policy Updates}

\title{Solving Consensus with OpenFlow}

\title{One-Shot Coordination of Distributed SDN Controllers:\\The Case for Multi-Write Consensus}

@@

\title{Coordinating Distributed SDN Controllers:\\On the Consensus Power of OpenFlow}

\author{
Liron Schiff\thanks{Supported by the European Research Council (ERC) Starting Grant no. 259085 and by the Israel Science Foundation Grant no.~1386/11.} $^1$,
Stefan Schmid$^2$, Petr Kuznetsov$^3$ \\
\small $^1$ Tel Aviv University, Israel; $^2$ TU Berlin \& T-Labs, Germany; $^3$ ParisTech, France
}

@@

\begin{abstract}
The control plane of modern Software-Defined Networks (SDNs)
is \emph{distributed}: a redundant control plane
can ensure a high availability, and a spatially distributed
control plane allows to handle time-critical dataplane
events close to their origins. Moreover, an SDN may 
be managed by multiple administrators simultaneously. 
However, not much is known today about how to implement
such a distributed control plane. 
In this paper, we show that powerful coordination objects
can be implemented using today's Openflow switches.
In particular, we describe the implementation of test\&set primitives 
as well as multi-writer consensus objects which can be used
to implement atomic transactions with all-or-nothing semantics.
Interestingly, our approach allows controllers to detect and
resolve very general notions of conflicts, not only, e.g., conflicts
due to overlapping flow spaces but also conflicts based on network load.
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller,
Software-defined networks (SDNs) simplify the network management and facilitate faster
innovations, as the (software) control plane is decoupled and can evolve independently.
Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
by relying on a simple match-action paradigm which allows us to define
forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond)
header fields.

While the logically centralized network perspective has the potential to simplify
the network management---many network tasks are inherently non-local---and introduces
many flexibilities,
managing and operating an SDN in a more adaptive but consistent manner is non-trivial, even with SDN.
A particularly interesting problem regards the consistent installation of new policies
or routes.~\cite{network-update,roger-hotnets,correct} While implementing consistent network updates
is challenging already from the perspective of a single controller, there exists a wide consensus
that the logically centralized SDN control planes will likely be \emph{actually distributed} in the future.

The reason for implementing SDN control planes in a distributed manner are manyfold.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently.~\cite{participatory,stn}

However, today, we do not have a good understanding of how to realize
such distributed control planes. The problem is essentially a distributed
systems one: Multiple controllers may simultaneously try to
install conflicting updates, e.g., rules (of the same priority) defined over non-disjoint
flow space, or load-balancing policies which, if combined, actually \emph{increase} the load~\cite{log-cent}. 
Moreover, controllers may also fail,
before their update has been completed.

\subsection{Our Contributions}

In this paper, we show that powerful synchronization primitives
can be implemented in Openflow, allowing controllers to implement
non-trivial atomic transactions, and detect and
automatically resolve conflicts. In particular, we show how to implement
the classic distributed computing primitive
\emph{test\&set}, supporting the conditional installation of new rules,
and we show how to use the bundle feature to implement
multi-writer objects, an efficient means to compute consensus;
interestingly, our implementation can also be used to compute a consensus
and install the winner's rule \emph{atomically}, using one message.
Moreover, our approach allows to detect and resolve very general notions of conflicts:
conflicts may be defined, for example, for rules of the same priority over
defined over non-independent flow spaces, but may also depend, e.g., on 
network load (based on counters).
We believe that our approach nicely complements
existing, more theoretical literature on distributed control planes; for example,
it also shows how to realize the atomic read-modify-write primitive postulated in
STN~\cite{stn}.

\subsection{Organization}

The remainder of this paper is organized as follows.
In Section~\ref{sec:background}, we provide the theoretical
background on our approach. Section~\ref{sec:realization}
then discusses the basic implementation of this approach in Openflow.
Section~\ref{sec:extensions} discusses practical issues and extensions.
After reviewing related literature in Section~\ref{sec:relwork},
we conclude our work in
Section~\ref{sec:conclusion}.

>>>>

%\usepackage[caption=true,font=footnotesize]{subfig}
%\usepackage{xspace} 	% Guesses whether a space is needed when invoked
\usepackage{cite}
\usepackage{url}

%\usepackage{framed}
\usepackage{framed,color}
\definecolor{shadecolor}{rgb}{0.9,0.9,0.9}

@@

%\title{Solving Consensus with OpenFlow}

%\title{One-Shot Coordination of Distributed SDN Controllers:\\The Case for Multi-Write Consensus}

@@

%\title{Coordinating Distributed SDN Controllers:\\On the Consensus Power of OpenFlow}

\title{On the Synchronization Power of OpenFlow}

\author{
Liron Schiff\thanks{Supported by the European Research Council (ERC) Starting Grant no. 259085 and by the Israel Science Foundation Grant no.~1386/11.} $^1$,
Stefan Schmid$^2$, Petr Kuznetsov$^3$ \\
\small $^1$ Tel Aviv University, Israel; $^2$ TU Berlin \& T-Labs,
Germany; $^3$ T\'el\'ecom ParisTech, France
}

@@

\begin{abstract}
Control planes of modern Software-Defined Networks (SDNs)
are typically \emph{distributed systems}: to achieve availability,
load balancing and responsiveness, control modules are installed on
multiple machines.   
%a redundant control plane
%can ensure a high availability, and a spatially distributed
%control plane allows to handle time-critical dataplane
%events close to their origins. 
%[[PK this is a contraversial point, not important for us 
%Moreover, an SDN may be managed by multiple administrators simultaneously. 
%]]
Unfortunately, these goals are hard to combine with consistent network
manipulation and certain synchronization challenges among the
distributed controllers must be addressed. 
%[[PK sounds contradicting to the fact that such systems already exist
%However, not much is known today about how to implement
%such a distributed control plane. 
%]]
In this paper, we show that powerful synchronization mechanisms
can be implemented \emph{in band}, simply using today's Openflow switches.
We show that standard OpenFlow features allow for providing simple
but general-purpose transactions with all-or-nothing semantics. 
We then show that these transactions can be used to reach consensus
among software controllers and, even more surprizingly, to resolve
\emph{on-the-fly} conflicts between concurrent policy updates in an
effectively atomic way, without affecting the ongoing traffic.  
%[[PK
%In particular, we describe the implementation of test\&set primitives 
%as well as multi-writer consensus objects which can be used
%to implement atomic transactions with all-or-nothing semantics.
%Interestingly, our approach allows controllers to detect and
%resolve very general notions of conflicts, not only, e.g., conflicts
%due to overlapping flow spaces but also conflicts based on network load.
%]]
\end{abstract}

@@

By consolidating and outsourcing the control over the dataplane switches to a logically
centralized controller, the concept Software-Defined Networking (SDN)
simplifies network management and facilitates faster innovations, 
as the (software) control plane is decoupled from the (hardware) data
plane and can evolve independently.
%[[PK do not see why it should be here
%Moreover, Openflow, the standard SDN protocol today, introduces many generalizations,
%in terms of traffic engineering, definition of flows, as well as in-band network functionalities,
%by relying on a simple match-action paradigm which allows us to define
%forwarding rules based not only on Layer-2, but also Layer-3 and Layer-4 (and possibly beyond)
%header fields.
%]]
%
The distinct network control plane has the potential to simplify
network management
%[[PK did not get
%---many network tasks are inherently non-local---
%]]
and opens new opportunities for innovation,
%[[PK we rather  talk about implementing it? 
%managing and operating an SDN in 
but implementing it 
%]]
%[[PK adaptive to what?
%a more adaptive but 
%]]
in a scalable and  consistent manner is highly non-trivial.
%[[PK did not get ``even''
%, even with SDN.
%]]

Indeed, there is a wide consensus that to achieve availability,
responsiveness, and balancing of network management tasks, 
the \emph{logically} centralized SDN
control plane must be physically \emph{distributed}.
First, in order to provide a high availability as well as a minimal degree of
fault-tolerance, controllers should be redundant~\cite{onix,stn,onos}: a failure
of one controller can be masked by other controllers. Second, it has also been proposed
to distribute controllers \emph{spatially}, in order to handle latency-sensitive and
communication intensive control plane events close to their origin.~\cite{devoflow,kandoo,jukka,disco}
Third, larger SDNs are likely to be operated by multiple administrators or may even offer
participatory interfaces where different users can install and trigger policy changes
concurrently~\cite{participatory,stn}.
All this, however, comes with the need for a \emph{synchronization}
between distributed controllers, assuming that we want them to
manipulate the network \emph{consistently}~\cite{cpc}.
And this task is highly nontrivial~\cite{cap-theorem}.

%[[PK not sure I see why talking about it here
%[[
%A particularly interesting problem regards the consistent installation of new policies
%or routes~\cite{network-update,roger-hotnets,correct,stn}. While implementing consistent network updates
%is challenging already from the perspective of a single controller, there exists a wide consensus
%that the logically centralized SDN control planes will likely be \emph{actually distributed} in the future.
%]]

Today, unfortunately, we do not have a good understanding yet of how to realize
such distributed control planes. The problem is essentially a
distributed-systems 
one: Multiple controllers may simultaneously try to
install conflicting updates and we want to resolve these conflicts
\emph{consistently} (no undesired behavior is observed on the data
plane) and \emph{efficiently} (no undesired delays are imposed on the
control application).

Consider, for example, the particularly interesting problem of
consistent installation of new forwarding policies, stipulating routes
that packets of different header spaces should follow across the
network~\cite{network-update,roger-hotnets,correct,stn}.
Installing conflicting forwarding rules, e.g., rules of the same priority defined over non-disjoint
flow spaces may lead to pathological network behavior (loops,
bleckholes, etc.).
Similarly, installing diverging load-balancing policies may, when combined, \emph{increase} the load~\cite{log-cent}. 
To render things more difficult, controllers may also fail, even before their updates have been completed.

%\subsection{In band synchronization?}
%
In this paper, we look closely at the most popular specification
language for SDN control-data plane exchange -
OpenFlow~\cite{of-spec}. 
At a high level, the specification represents the exchange in the form
of \emph{control messages}, containing commands for installing or
modifying \emph{rules} that, in turn, constitute the \emph{flow table} of
a data-plane switch used in processing the ongoing data-plane traffic.

We observe that the standard provides the option of \emph{bundling}
these messages in one, so that either all of the control messages are
processed or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.       
Interestingly, this bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.

Intuitively, if we treat the flow table at the switch as a \emph{shared memory}
that can be concurrently read and modified by multiple controllers, 
the bundling features can be seen as a \emph{multi-write} or, more
widely, \emph{multi-modify} operation that modifies
multiple memory location in one atomic step.
This ability is reminiscent of the hypothetical multi-write
shared-memory model which is known to enable solutions for important
synchronization problems, such as consensus, impossible to solve
otherwise~\cite{multi-objects}. 
Indeed, we describe an algorithm that solves consensus among any
number of controllers using a single OpenFlow switch that accepts
bundle control messages.

Even more, we observe that the standard also assumes \emph{conditional} application of
certain types of control messages. A specific $\textsf{OFPFF}\_\textsf{CHECK}\_\textsf{OVERLAP}$ flag
set in the control message containing a new forwarding rule ensures
that the rule is applied if and only if it does not \emph{conflict}
with any other rule currently in the flow table.  
This in-band conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.

We then discuss how these synchronization mechanisms can be used to
achieve consistent network management. 
We first show that they can be used to replace the read-modify-write
primitives voluntarily assumed in the recent implementation of a
recent proposal for a consistent composition system for forwarding policies~\cite{cpc}.  
We then use these transactions directly to implement a lightweight
forwarding policy composition system that requires no additional
synchronization on the control plane.     
We then explore how more general classes of policies can be composed
\emph{on-the-switch}.

% we show that powerful synchronization primitives
%can be implemented in Openflow, allowing controllers to implement
%non-trivial atomic transactions, and detect and
%automatically resolve conflicts. In particular, we show how to implement
%the classic distributed computing primitive
%\emph{test\&set}, supporting the conditional installation of new rules,
%and we show how to use the bundle feature to implement
%multi-writer objects, an efficient means to compute consensus;
%interestingly, our implementation can also be used to compute a consensus
%and install the winner's rule \emph{atomically}, using one message.
%Moreover, our approach allows to detect and resolve very general notions of conflicts:
%conflicts may be defined, for example, for rules of the same priority over
%defined over non-independent flow spaces, but may also depend, e.g., on 
%network load (based on counters).
%We believe that our approach nicely complements
%existing, more theoretical literature on distributed control planes; for example,
%it also shows how to realize the atomic read-modify-write primitive postulated in
%STN~\cite{stn}.

%\paragraph{Organization.}
%
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we provide the theoretical
background on our approach. Section~\ref{sec:realization}
then discusses the basic implementation of this approach in Openflow.
Section~\ref{sec:extensions} discusses practical issues and extensions.
After reviewing related literature in Section~\ref{sec:relwork},
we conclude our work in
Section~\ref{sec:conclusion}.
